---
title: "The basic workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{import_gc_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

My colleagues and I follow a similar, if not the same, methodology for the 
analysis of cuticular hydrocarbons (CHCs), considering the overall description 
of the methods (e.g.extraction and Gas chromatography-Mass spectrometry (GC-MS)
analysis of these compounds). However, we have noticed that for the data 
processing and analysis (statistics aside), we all have developed different
approaches/workflows, which are not always (entirely) reproducible. In fact, 
they sometimes lead to confusion among us, regarding how exactly the data sets 
of each other were/are being built. The so-called "master tables" (our final 
data sets), as well as the steps to produce them, end up having structures that
very much depend on each individual, regardless of the data, the statistical
analyses that will be applied to them, or the objective of the study they were
produced for.

Furthermore, every time that we have to explain to a student how to process the
data to build up the master table and analyze the resulting data sets, the
student has to face the confusion delivered by the incompatibilities/differences
of the different steps of the processes from each of us. Besides, each of us has
to make an excessive time investment, by trying to connect the different steps
and methods, in order to explain them to the students (or anybody else).

Therefore, I decided to build up this package to facilitate the process of
analyzing this and similar types of GC data in R, thus providing a tool set and
reference for a semi-automated and reproducible workflow.

This guide offers a step-by-step explanation of such workflow, focusing on the
analysis of GC-MS data of cuticular hydrocarbons. That said, my intention here 
is not to teach (nor show) how to extract CHCs, perform a GC-MS run with such
extracts, or to interpret the resulting chromatograms and mass-spectra to
assess their composition.

Assuming that you already have an idea to do this, this guides start from the 
point at which we would have already run a all our samples (extracts) in the
GC-MS machine, and integrated their resulting total ion count (TIC) 
chromatograms. Therefore, we have a collection of tabular files (i.e. CSV)
containing these results.

Now, let us begin by loading the package in our R session, ot be able to access
its functions.

```{r setup}
library(analyzeGC)
```

# Import the GC-MS integration data


